---
format: gfm
editor: visual
---

```{r}
#| echo: true
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
```

## Exercise 1: Unit converter

**1.1 Write a function that can convert Fahrenheit to Celsius**, based
on the following fomula: `C = (F - 32) * 5 / 9`

```{r}
#| echo: true
#| message: false
#| warning: false
f_to_c <- function(f) {
  c <- (f - 32) * 5 / 9
  return(c)
}
```

Take your function for a spin, does it return the correct values?

32 F = 0 C

```{r comment="##"}
#| echo: true
#| message: false
#| warning: false
f_to_c(32)
```
50 F = 10 C

```{r comment="##"}
#| echo: true
#| message: false
#| warning: false
f_to_c(50)
```

212 F = 100 C

```{r comment="##"}
#| echo: true
#| message: false
#| warning: false
f_to_c(212)
```

<br>

**1.2** A student came from ‘tropical Canada’. She doesn’t like the cold
but she really didn’t like it when it’s hot. Although she wanted to know
what the temperature is in Celsius when the US weather channel reported
it in Fahrenheit, there are certain points at which it was just too cold
or too hot for to to care about the exact value. **Modify the f_to_c
function below to print the following**, and check if your function
works properly using the input of **-10 F, 60 F, and 90 F**.

- If the temperature is less than -20 C, print “Don’t bother going out.”

- If the temperature is greater than 30 C, print “I’m moving back to
  Canada.”

- If the temperature is between -20 C and 30 C, print the temperature in
  Celsius.

```{r}
#| echo: true
#| message: false
#| warning: false
f_to_c_message <- function(f) {
  c <- (f - 32) * 5 / 9
  
  if (c < -20) {
    print("Don't bother going out.")
  }
  else if (c > 30) {
    print("I'm moving back to Canada.")
  }
  else {
     return(c) 
  }
}
```

The following is how it should work:

```{r comment="##"}
f_to_c_message(-10)
f_to_c_message(60)
f_to_c_message(90)
```

<br> <br>

## Exercise 2: Set operation

The R function `setdiff(x,y)` returns the elements of a vector x that
are not in y. For example, say x is a list of singers who can reach very
high registers, and y is a list of musicians who died in 2016.
`setdiff(x, y)` would return a list of singers with high voices who did
not die in 2016.

```{r comment="##"}
#| echo: true
#| message: false
#| warning: false
x <- c("prince", "mj", "sam cook", "whitney", "dolly")
y <- c("sharon jones", "prince", "bowie", "leonard cohen", "phife dawg")
setdiff(x,y)
```

However, sometimes you want a function to return the elements that are
in x or y but not both. Here, **write such a function**, and use your
custom function to **find the names of singers who either have high
voices, but did not die in 2016, OR who died in 2016 but do not have
high voices.**

Your function will need to work for any vectors, not just the ones in
this example. The output should be a single vector, not two vectors.

**Hint**: you will need R’s [set
operations](https://stat.ethz.ch/R-manual/R-devel/library/base/html/sets.html).

```{r}
#| echo: true
#| message: false
#| warning: false
unique_element <- function(x,y){
  only_in_x <- setdiff(x,y)
  only_in_y <- setdiff(y,x)
  result <- c(only_in_x, only_in_y)
  return(result)
}
```


The following is how it should work:

```{r comment="##"}
#| echo: true
#| message: false
#| warning: false
#Example output
unique_element(x,y)
```

<br> <br>

## Exercise 3: Fuel Efficiency of Car Models

This exercise uses the `mtcars` dataset, which is available in base R.
It was extracted from the 1974 Motor Trend US magazine, and comprises
fuel consumption and 10 aspects of automobile design and performance for
32 automobiles (1973–74 models). Let’s first convert `mtcars` to tibble
format such that the row names become a column named “model”.

```{r}
#| echo: true
#| message: false
#| warning: false
mtcars_tbl <- as_tibble(mtcars, rownames = "model")
mtcars_tbl %>%
  head() %>%
  kable()
```

From `mtcars_tbl`, reproduce the following plot, which shows the miles
per gallon (`mpg`) of car models on the x axis (see hints below).
Different models are ordered on the y axis according to their `mpg` and
their names are shown next to the data points. Also, the size of each
data point maps to its horse power (`hp`), and the color maps to number
of cylinders (`cyl`).

<br>

**Hints:**

- Start by reordering the factor levels of the `model` column based on
  values in the `mpg` column.

- You may also need to convert `cyl` first into a character variable
  (`as.character()`) or a factor (`factor` or `as_factor()`), and then
  (re)set its factor levels.

- When making the plot, please pay attention to the following elements
  of the plot that need to be specified.

  - Horizontal adjustment of the text labels (“`hjust`” option in
    `geom_text()`)
  - The range of the x-axis (`xlim()`)
  - x-axis label and legend titles (`labs()`)
  - The preset theme (`theme_*()`)
  - The removal of the y-axis (`theme()`)

<br>

```{r}
#| echo: true
#| message: false
#| warning: false
mtcars_plot <- mtcars_tbl |> 
  mutate(model = fct_reorder(model, mpg),
         cyl = fct_rev(as.factor(cyl))) |> 
  ggplot(aes(x = mpg, y = model, size = hp, color = cyl)) +
  geom_point() +
  geom_text(aes(label = model), hjust = -0.1, color = "black", show.legend = F, size = 4) +
  xlim(10, 40) +
  labs(x = "Miles per gallon fuel consumption",
       y = NULL,
       size = "Horsepower",
       color = "Number of cylinders") +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank()) +
   guides(size = guide_legend(order = 1),
         color = guide_legend(order = 2))
mtcars_plot
```

<br>

#### **Reflection Questions:**

<br>

#### In this assignment, you wrote functions that operate on vectors (Exercises 1 and 2) and manipulated a data frame to create a plot (Exercise 3). Explain the difference between a simple character vector (like the one for singers) and a factor (like the model column). Why was it necessary to treat the model column as a factor and reorder its levels to create the final plot?

A character vector is a sequence of character strings, while a factor is a data type that represents categorical data. It was necessary to treat the model column as a factor and reorder its levels to create the final plot so that the y-axis could be arranged from highest to lowest fuel effeciency. If I had not done this, then it would have been arranged in alphabetical order.

<br>

#### Identify one of the three exercises where you used an AI assistant to help generate or debug a chunk of code. Provide the specific prompt you used. Then, show the code before you finalized it (e.g., the AI's initial suggestion or your first attempt) and the final, working code. In your commentary, explain what was missing or incorrect in the "before" version and what you learned by testing and correcting it to produce the final version.

When I was working on problem 3, I asked Claude this: 

"mtcars_plot <- mtcars_tbl |> mutate(model = fct_reorder(model, mpg), cyl = as.factor(cyl)) |> ggplot(aes(x = mpg, y = model, size = hp, color = cyl)) + geom_point() + geom_text(aes(label = model), hjust = -0.1) + labs(x = "Miles per gallon fuel consumption", y = NULL, size = "Horsepower", color = "Number of cylinders") mtcars_plot 

I want the number of cylinders to be in reverse order" 


This is the code for my question:

```{r}
#| echo: true
#| message: false
#| warning: false
mtcars_plot_ai1 <- mtcars_tbl |> 
  mutate(model = fct_reorder(model, mpg),
         cyl = as.factor(cyl)) |> 
  ggplot(aes(x = mpg, y = model, size = hp, color = cyl)) +
  geom_point() +
  geom_text(aes(label = model), hjust = -0.1) +
  labs(x = "Miles per gallon fuel consumption",
       y = NULL,
       size = "Horsepower",
       color = "Number of cylinders")
mtcars_plot_ai1
```
Claude said that I can use fct_rev() when I am setting cyl to a factor (3rd line of code in that chunk).

```{r}
#| echo: true
#| message: false
#| warning: false
# This is the code Claude gave me

mtcars_plot <- mtcars_tbl |> 
  mutate(model = fct_reorder(model, mpg),
         cyl = fct_rev(as.factor(cyl))) |> 
  ggplot(aes(x = mpg, y = model, size = hp, color = cyl)) +
  geom_point() +
  geom_text(aes(label = model), hjust = -0.1) +
  labs(x = "Miles per gallon fuel consumption",
       y = NULL,
       size = "Horsepower",
       color = "Number of cylinders")
mtcars_plot
```

This new code did what I expected, which was it essentially reversed the color order of the graph and the order of the key (I wanted my graph to match the graph in the assignment.)

<br>